---
title: "Estimate Response to Selection"
author: "Andrew Morris"
date: "`r Sys.Date()`"
output: html_document
---

This script intakes the flux data from script 01. It then calculates deviance
as the difference between the positive line and the neutral line in each
passage. The errors can either be calculated as the sum of the errors for
each line or the standard error of difference, that is, the square root of
the sum of the squared errors, which is slightly smaller than the sum. In
this case, I have chosen the SED.
```{r initialize_packages, message=FALSE}
library(tidyverse)
library(morris)
library(broom)
library(knitr)
```

```{r initialize_variables}
source('functions.R')
```

```{r import_data, message=FALSE}
fluxes <- read_tsv(paste0(der_dir, 'fluxes.tsv'))
```

## Plot the raw data

```{r calc_plot_ratios}

```

```{r plot_raw}
ratio <- calc_plot_ratios(fluxes$passage, fluxes$estimate)

plot_flux(fluxes, passage, estimate, ratio)
#ggsave('fluxes_raw.pdf', width = 5, height = 4)
```

## Plot the log-transformed data

```{r plot_log10}
ratio_log10 <- calc_plot_ratios(fluxes$passage, fluxes$estimate_log10)

plot_flux(fluxes, passage, estimate_log10, ratio_log10, log10 = TRUE)

#ggsave('fluxes_log.pdf', width = 5, height = 4)
```

## Test a difference of slopes

```{r response_model}
# Fit full interaction model
inter_model <- lm(estimate_log10 ~ 0 + passage * treat, data = fluxes)

# Fit additive model
addit_model <- lm(estimate_log10 ~ passage + treat, data = fluxes)


# Back calculate the slope of the positive treatment
pos_slope <- 10^inter_model$coef["passage:treatp"]
```

Here I fit two models with flux as the dependent variable and passage and treatment as the independent variables. The first is the interaction model, which includes an interaction between passage and treatment. The second is an additive model that lacks this interaction. Here is the model comparison using `anova`, which indicates the full model is somewhat better:

```{r compare_models}
# Compare interaction to no interaction
compare_mods <- anova(inter_model, addit_model)
kable(tidy(compare_mods))
```

Here is the full interaction model:

```{r inter_model}
kable(tidy(inter_model))
```

The term `passage` was not significant, which is the y-intercept of the neutral treatment So This treatment did not change. 

## Previous Method

Previously, I was using the `metafor` package to test the slope of the difference between the positive and neutral line. 

```{r eval=FALSE}

library(metafor)
# Standard error of the difference
sed <- function(p, n, na.rm = FALSE) {
  sqrt(se(p, na.rm = na.rm) ^ 2 + se(n, na.rm = na.rm)^ 2)
}
# Calculate deviance as p - n
deviance <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  group_by(passage, treat) %>% 
  summarize(mean = mean(estimate), .groups = "drop") %>% 
  pivot_wider(names_from = treat, values_from = mean) %>% 
  mutate(deviance = p - n)

# Calculate variance of deviance as SED
deviance$se <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  mutate(x = 1:nrow(fluxes)) %>% 
  pivot_wider(names_from = treat, values_from = estimate) %>% 
  group_by(passage) %>% 
  summarize(sed = sed(p, n, na.rm = TRUE), .groups = "drop") %>%
  pull(sed)

# Fit deviance model
fit <- rma(deviance ~ passage, se^2, method="FE", data = deviance)

plot(fit)

# Plot deviance
ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_pointrange(color = 'darkorange2') + 
  geom_abline(intercept = fit$beta[1], slope = fit$beta[2])

# Take deviance model parameters and export into a data.frame
rma_output <- data.frame(
Estimate = fit$beta,
se = fit$se,
z.value = fit$zval,
p.value = fit$pval,
upper.ci = fit$ci.lb,
lower.ci = fit$ci.ub
)

# Fit deviance model without passage 2
fit_no2 <- 
filter(deviance, passage != 2) %>% 
rma(deviance ~ passage, se^2, method="FE", data = .)

# Take deviance model withou passage 2 parameters and export into a data.frame

rma_output_no2 <- data.frame(
Estimate = fit_no2$beta,
se = fit_no2$se,
z.value = fit_no2$zval,
p.value = fit_no2$pval,
upper.ci = fit_no2$ci.lb,
lower.ci = fit_no2$ci.ub
)

# R (response to selection) for the model w/o sample 2
R_no2 <- rma_output_no2[[2, 1]]
R <- rma_output[[2, 1]]

# Plot the no2 model
filter(deviance, passage != 2) %>% 
ggplot(., aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_pointrange(color = 'darkorange2') + 
  geom_abline(intercept = fit_no2$beta[1], slope = fit_no2$beta[2])

#write_tsv(deviance, '../Output/deviance.tsv')
#saveRDS(rma_output, '../Output/response.rds')
#saveRDS(fit, '../Output/dev_fit.rds')



# Calculate deviance as percentage of p - n
deviance <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  group_by(passage, treat) %>% 
  summarize(mean = mean(estimate), .groups = "drop") %>% 
  pivot_wider(names_from = treat, values_from = mean) %>% 
  mutate(deviance = (p - n) / n)

# Calculate variance of deviance as SED
deviance$se <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  mutate(x = 1:nrow(fluxes)) %>% 
  pivot_wider(names_from = treat, values_from = estimate) %>% 
  group_by(passage) %>% 
  summarize(sed = sed(p, n, na.rm = TRUE), .groups = "drop") %>%
  pull(sed)

# Fit deviance model
fit <- rma(deviance ~ passage, se^2, method="FE", data = deviance)

plot(fit)

# Plot deviance
ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_pointrange(color = 'darkorange2') + 
  geom_abline(intercept = fit$beta[1], slope = fit$beta[2])

# Take deviance model parameters and export into a data.frame
rma_output <- data.frame(
Estimate = fit$beta,
se = fit$se,
z.value = fit$zval,
p.value = fit$pval,
upper.ci = fit$ci.lb,
lower.ci = fit$ci.ub
)

# Fit separte slopes for Positive and Neutral with robust M-estimator regression

library(MASS)
anova(
fit <- rlm(estimate ~ passage * treat, data = fluxes),
fit <- rlm(estimate ~ passage + treat, data = fluxes))

summary(fit)
anova(fit)
new_data <- cbind(fluxes, predict(fit, interval = 'confidence'))
colnames(new_data) 
ggplot(new_data, aes(x = passage, y = estimate, color = treat)) +
  theme_bw() +
  geom_jitter() +
  geom_line(aes(passage, fit)) +
  geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)

ggplot(new_data, aes(x = passage, y = estimate)) +
  theme_bw() +
  geom_jitter(aes(color = treat)) +
  geom_line(aes(passage, fit, color = treat)) +
  geom_ribbon(aes(ymin=lwr,ymax=upr, group = treat), alpha=0.3) +
  ylim(-0.01, 0.08)
ggplot(fluxes, aes(x = passage, y = estimate, color = treat)) + 
  geom_point() + 
  stat_smooth(method='lm')
#ggsave('raw_fluxes.pdf')
```